Beyond Ultra-Diffuse Galaxies I: Mass-Size Outliers Among the Satellites of Milky Way Analogs

### 2023-06-02 paper1.tex
shared with coauthors

The layout of this paper is as follows. Section \ref{sec:data} describes the data used in this work. In Section \ref{sec:lsbg_search}, we describe the LSBG search in HSC data, including source detection (\S \ref{sec:detection}), deblending (\S \ref{sec:deblending}), modeling (\S \ref{sec:modeling}), completeness, and uncertainty (\S\ref{sec:comp_meas}). In Section \ref{sec:sample_construction}, we propose a new definition for large and diffuse galaxies (the UPGs) and present our UDG and UPG samples. In Section \ref{sec:results}, we compare the UDGs and UPGs on the mass-size plane. We calculate their abundances and compare with the literature (\S \ref{sec:n_udg}). Based on these results, we further argue that we should move beyond UDGs and adopt the UPG definition in Section \ref{sec:discussion}. Section \ref{sec:summary} presents a summary of this work and prospects for the future. In a companion paper \citep{Li2023}, we discuss the size distributions, spatial distributions, and quenching of UDGs and UPGs.

@@

Continuing the work of \citet{Greco2018} (hereafter \citetalias{Greco2018}), we conduct a systematic search for low surface brightness galaxies in the HSC PDR2 data, which covers $\sim 1.5$ times the sky area of \citetalias{Greco2018}. The schematic of the searching method in this paper remains similar to that in \citetalias{Greco2018}, but we do improve the search completeness and purity by adding several new steps, which are highlighted in this section. We outline the major steps below and refer the reader who are already familiar with the algorithms in LSBG searches to \S \ref{sec:sample_construction} for the mass-size outlier sample construction.

@@

In \code{scarlet}, each source $k$ in the cutout is described by a morphology image $S_k$ and a Spectral Energy Distribution (SED) vector $A_k$. The multi-band images are represented as $Y$. The goal of the modeling is to minimize the objective function $f = \frac{1}{2} ||Y - P \ast (\sum_k A_k^\top \times S_k)||^{2}$ under certain constraints, where $P$ is the PSF and $*$ is convolution. The morphology image of each source is limited by a bounding box. We assume two constraints when running \code{scarlet}: all sources have positive fluxes (positivity constraint) and the light profiles of all sources monotonically decrease from center to outskirts (monotonicity constraint). The monotonicity constraint is over-simplified for well-resolved galaxies with complicated structures, but for objects in our sample this assumption still holds for most cases and provides an effective and robust way to deblend overlapping sources. We refer to this modeling method as \textit{vanilla \code{scarlet}}.

@@

One of the most important diagnostics is the effective radius of the object. In this paper, we refer to effective radius as the circularized effective radius $r_{e}$, defined as the $r_{e} = r_{\rm eff, sma} \sqrt{b/a}$ where $r_{\rm eff, sma}$ is the effective radius measured along the semi-major axis of the aligned elliptical isophotes, and $b/a$ is the axis ratio of the isophotes. In \code{statmorph}, the effective radius is calculated using elliptical apertures, where the ellipticity $\varepsilon$ is determined by calculating the second moment of the image. The total magnitudes are simply calculated by summing up the model flux in each band, and colors are defined using total magnitudes.\footnote{A Galactic extinction correction is not applied at this point.} The central surface brightness $\mu_0$ is measured by \emphasize{linearly extrapolating the surface brightness profile to $r\to 0$ using \code{scipy} function \footnote{\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.interpolate.interp1d.html}} \citep{scipy}.} We define the average surface brightness within the effective radius as 
\begin{equation}\label{eq:mu_eff}
    \overline{\mu}_{\rm eff} = m - 2.5 \log_{10}(2 \pi r_e^2),
\end{equation}
where $m$ is the total magnitude.

@@

\jiaxuan{Need to clearly cite \citet{Lim2020}, also cite \citet{LaMarca2022,Venhola2022}. also \citep{Trujillo2017,ManceraPina2018} this issue wont manifest since every UDG is quenched in clusters.}

@@

\begin{align*}
&\log \left(M_{\star} / L_{g}\right)=1.774\,(g-r)-0.783, \\
&\log \left(M_{\star} / L_{g}\right)=1.297\,(g-i)-0.855.
\end{align*}

Because we have both $g-r$ and $g-i$ colors available from the model fitting, we use the average $M_{\star}/L$ derived from the two colors to calculate stellar mass. We assume the solar absolute magnitude in $g$-band to be 5.03 \citep{Willmer2018}. \emphasize{We note that the stellar mass derived using the color-$M_\star/L$ relation from \citet{Herrmann2016,Du2020} is $\sim 0.2$ dex lower than the one derived using \citet{Into2013} relation.}

@@

We highlight the similarities and differences between the UDGs and UPGs in the right panel. UDGs that are also classified as UPGs are shown as orange squares, whereas UDGs that do not satisfy the UPG definition are marked as red stars. Blue triangles correspond to UPGs that are not classified as UDGs. Limited by the detection limit and the fact that our hosts lie beyond the Local Volume, our mass limit is $M_\star \approx 10^{6.5}\ M_\odot$, which is $\sim 1$ dex higher than ELVES. 
Compared with the UPG sample, we find that the UDG sample (orange squares + red stars) comprises a number of galaxies below the $1.5\sigma$ line, but also loses several low-mass galaxies that are above the $1.5\sigma$ line but have sizes smaller than 1.5 kpc. For the UDG sample, the stellar mass range only reaches $M_\star\approx 10^{8.3}\ M_\odot$ because galaxies with higher stellar mass are too bright to satisfy the UDG surface brightness criterion. On the contrary, since the UPG definition does not have a hard surface brightness cut, it includes many galaxies that are more massive than UDGs provided they have an extraordinarily large size. We also note that the mass-size relation from \citet{ELVES-I} is extrapolated to define UPGs in $10^{8.5}\ M_\odot < M_\star < 10^9\ M_\odot$, where about 10\% our UPGs are located (dotted green lines in Figure \ref{fig:mass_size}). However, the mass-size relation might show color and morphology dependence for $M_\odot > 10^{8.5}\ M_\odot$. \emphasize{As a consequence, disk galaxies might contribute to the UPG sample at the high-mass end. We visually checked the UPG sample and do find some blue UPGs resembling disk galaxies at the high-mass end. Spectroscopic follow-ups are needed to confirm whether they are just blue background galaxies.}

@@

@@

For the deblending completeness, we inject 5,000 mock \sersic{} galaxies into the coadd and run vanilla \code{scarlet} on them. The mock galaxies follow the same uniform distribution in size, surface brightness, ellipticity, \sersic{} index as for deriving the detection completeness, but follow a Gaussian distribution in color: $g-i \sim \mathcal{N}(0.6, 0.2^2),\ g-r = 0.7 \cdot (g-i) + \mathcal{N}(0, 0.03^2)$. \emphasize{This mock galaxy population represents the real UDG/UPG population well. To be specific, the Gaussian color prior covers 86\% of the UPGs within $1.5\sigma$, the ellipticity prior covers 98\% of them, and the \sersic{} index prior covers 60\% of them.}
We measure the size and surface brightness on the scarlet models of the mock galaxies, then we apply the deblending cuts. We also find that the deblending cut mainly depends on the size and surface brightness.

>>>>

The layout of this paper is as follows. Section \ref{sec:data} describes the data used in this work. In Section \ref{sec:lsbg_search}, we describe the LSBG search in HSC data, including source detection (\S \ref{sec:detection}), deblending (\S \ref{sec:deblending}), modeling (\S \ref{sec:modeling}), completeness, and uncertainty (\S\ref{sec:comp_meas}). In Section \ref{sec:sample_construction}, we propose a new definition for large and diffuse galaxies (the UPGs) and present our UDG and UPG samples. In Section \ref{sec:results}, we compare the UDGs and UPGs on the mass-size plane. We calculate their abundances and compare them with the literature (\S \ref{sec:n_udg}). Based on these results, we further argue that we should move beyond UDGs and adopt the UPG definition in Section \ref{sec:discussion}. Section \ref{sec:summary} presents a summary of this work and prospects for the future. In a companion paper \citep{Li2023}, we discuss the size distributions, spatial distributions, and quenching of UDGs and UPGs.

@@

Continuing the work of \citet{Greco2018} (hereafter \citetalias{Greco2018}), we conduct a systematic search for low surface brightness galaxies in the HSC PDR2 data, which covers $\sim 1.5$ times the sky area of \citetalias{Greco2018}. The schematic of the searching method in this paper remains similar to that in \citetalias{Greco2018}, but we do improve the search completeness and purity by adding several new steps, which are highlighted in this section. We outline the major steps below and refer the readers who are already familiar with the algorithms in LSBG searches to \S \ref{sec:sample_construction} for the mass-size outlier sample construction.

@@

In \code{scarlet}, each source $k$ in the cutout is described by a morphology image $S_k$ and a Spectral Energy Distribution (SED) vector $A_k$. The multi-band images are represented as $Y$. The goal of the modeling is to minimize the objective function $f = \frac{1}{2} ||Y - P \ast (\sum_k A_k^\top \times S_k)||^{2}$ under certain constraints, where $P$ is the PSF and $*$ is convolution. The morphology image of each source is limited by a bounding box. We assume two constraints when running \code{scarlet}: all sources have positive fluxes (positivity constraint) and the light profiles of all sources monotonically decrease from the center to the outskirts (monotonicity constraint). The monotonicity constraint is over-simplified for well-resolved galaxies with complicated structures, but for objects in our sample this assumption still holds for most cases and provides an effective and robust way to deblend overlapping sources. We refer to this modeling method as \textit{vanilla \code{scarlet}}.

@@

One of the most important diagnostics is the effective radius of the object. In this paper, we refer to the effective radius as the circularized effective radius $r_{e}$, defined as the $r_{e} = r_{\rm eff, sma} \sqrt{b/a}$ where $r_{\rm eff, sma}$ is the effective radius measured along the semi-major axis of the aligned elliptical isophotes, and $b/a$ is the axis ratio of the isophotes. \footnote{\emphasize{We note that some authors use $r_{\rm eff, sma}$ as effective radius. There is no consensus on which choice is better in the literature.}} In \code{statmorph}, the effective radius is calculated using elliptical apertures, where the ellipticity $\varepsilon$ is determined by calculating the second moment of the image. The total magnitudes are simply calculated by summing up the model flux in each band, and colors are defined using total magnitudes.\footnote{A Galactic extinction correction is not applied at this point.} The central surface brightness $\mu_0$ is measured by \emphasize{linearly extrapolating the surface brightness profile to $r\to 0$ using \code{scipy} function\footnote{\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.interpolate.interp1d.html}} \citep{scipy}.} We define the average surface brightness within the effective radius as 
\begin{equation}\label{eq:mu_eff}
    \overline{\mu}_{\rm eff} = m - 2.5 \log_{10}(2 \pi r_e^2),
\end{equation}
where $m$ is the total magnitude.

@@

@@

\begin{align*}
&\log \left(M_{\star} / L_{g}\right)=1.774\,(g-r)-0.783, \\
&\log \left(M_{\star} / L_{g}\right)=1.297\,(g-i)-0.855.
\end{align*}
\emphasize{\citet{Kado-Fong2022} show that the UDG population can be described by the color-$M_\star/L$ relation from \citet{Into2013}. The stellar mass of dwarf in the ELVES survey is also derived using this relation \citep{CarlstenELVES2022}.} In our case, because we have both $g-r$ and $g-i$ colors available from the model fitting, we use the average $M_{\star}/L$ derived from the two colors to calculate stellar mass. We assume the solar absolute magnitude in $g$-band to be 5.03 \citep{Willmer2018}.\footnote{\emphasize{We note that the stellar mass derived using the color-$M_\star/L$ relation from \citet{Herrmann2016} and \citet{Du2020} is $\sim 0.2$ dex lower than the one derived using \citet{Into2013} relation. We use \citet{Into2013} prescription in this work encouraged by the finding in \citet{Kado-Fong2022} and for being consistent with ELVES.}}

@@

We highlight the similarities and differences between the UDGs and UPGs in the right panel. UDGs that are also classified as UPGs are shown as orange squares, whereas UDGs that do not satisfy the UPG definition are marked as red stars. Blue triangles correspond to UPGs that are not classified as UDGs. Limited by the detection limit and the fact that our hosts lie beyond the Local Volume, our mass limit is $M_\star \approx 10^{6.5}\ M_\odot$, which is $\sim 1$ dex higher than ELVES. 
Compared with the UPG sample, we find that the UDG sample (orange squares + red stars) comprises a number of galaxies below the $1.5\sigma$ line, but also loses several low-mass galaxies that are above the $1.5\sigma$ line but have sizes smaller than 1.5 kpc. For the UDG sample, the stellar mass range only reaches $M_\star\approx 10^{8.3}\ M_\odot$ because galaxies with higher stellar mass are too bright to satisfy the UDG surface brightness criterion. On the contrary, since the UPG definition does not have a hard surface brightness cut, it includes many galaxies that are more massive than UDGs provided they have an extraordinarily large size. We also note that the mass-size relation from \citet{ELVES-I} is extrapolated to define UPGs in $10^{8.5}\ M_\odot < M_\star < 10^9\ M_\odot$, where about 10\% our UPGs are located (dotted green lines in Figure \ref{fig:mass_size}). However, the mass-size relation might show color and morphology dependence for $M_\odot > 10^{8.5}\ M_\odot$. \emphasize{As a consequence, disk galaxies might contribute to the UPG sample at the high-mass end. We visually checked the UPG sample and do find some blue UPGs resembling disk galaxies at the high-mass end. Spectroscopic follow-ups are needed to confirm whether they are just blue background galaxies.} \emphasize{We also note that 7\% of UDGs and 3\% of UPGs have \sersic{} indices $n>1.5$. We believe these galaxies with large \sersic{} indices are mostly interlopers given the fact that $\sim$40\% of UDGs and UPGs are interlopers (see \S\ref{sec:bkg}). We do not consider this as an issue given their small contribution to the whole sample.}

@@

\begin{figure*}
	\vbox{ 
		\centering
		\includegraphics[width=0.9\linewidth]{figures/mock_sample_upg.pdf}
	}
    \caption{The distribution of UPGs (blue) and mock galaxies (red) in $g-i$ color, ellipticity, and \sersic{} index $n$. We translate the Spergel index to the \sersic{} index using the relation \eqref{eq:symbolic}. The mock galaxy population represents the real UDG/UPG population well.
    }
    \label{fig:mock_params}
\end{figure*}

@@

For the deblending completeness, we inject 5,000 mock \sersic{} galaxies into the coadd and run vanilla \code{scarlet} on them. The mock galaxies follow the same uniform distribution in size, surface brightness, ellipticity, \sersic{} index as for deriving the detection completeness, but follow a Gaussian distribution in color: $g-i \sim \mathcal{N}(0.6, 0.2^2),\ g-r = 0.7 \cdot (g-i) + \mathcal{N}(0, 0.03^2)$. \emphasize{We show the distribution of mock galaxies and UPGs in Figure \ref{fig:mock_params}. This mock galaxy population represents the real UDG/UPG population well. To be specific, the Gaussian color prior covers 86\% of the UPGs within $1.5\sigma$, the ellipticity prior covers 98\% of them, and the \sersic{} index prior covers 60\% of them.}
We measure the size and surface brightness on the scarlet models of the mock galaxies, then we apply the deblending cuts. We also find that the deblending cut mainly depends on the size and surface brightness.
